---
title: "John Hopkins Practical Machine Learning Report"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
The project for this course is to use machine learning to predict the manner in which a group did excercise. The following objectives are to be met as part of this report:
* Describe how you built your model
* How you used cross validation
* Why you made the choices you did

### Project Setup
First, include required libraries, create folders, and download files

```{r libraries, cache=TRUE, message=FALSE}
# Include libraries
library(caret)
```

```{r folders, cache=TRUE}
# Create data folder if it does not exist
dataFolder = "data"
if (!file.exists(dataFolder)) {
        dir.create(dataFolder)
}
```

```{r files, cache=TRUE}
# Download test files if they do not exist
trainInput <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testInput <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
trainDisk <- file.path(dataFolder,"training.csv")
testDisk <- file.path(dataFolder, "test.csv")

if (!file.exists(trainDisk)) {
        download.file(trainInput, destfile = trainDisk, method = "curl")
}

if (!file.exists(testDisk)) {
        download.file(testInput, destfile = testDisk, method = "curl")
}
```

### Prepare Data
Next we need to load, clean, and partition our data sets

```{r load, cache=TRUE}
# Read the downloaded csv files into data frames
trainCsv <- read.csv(trainDisk)
testCsv <- read.csv(testDisk)

# Next I used View(trainCsv) and I see lots of NA Values. Lets remove those first.
trainCsv <- trainCsv[, colSums(is.na(trainCsv)) == 0]
testCsv <-  testCsv[, colSums(is.na(testCsv)) == 0]

# Again I used View(trainCsv) and there are some obvious columns that are not useful
dropCol <- c("user_name", "raw_timestamp_part_1", "raw_timestamp_part_2", "cvtd_timestamp", "new_window", "num_window")
trainCsv <- trainCsv[, -which(names(trainCsv) %in% dropCol)]
testCsv <- testCsv[, -which(names(testCsv) %in% dropCol)]

# Finally partition the data
set.seed(19622)
inTrain <- createDataPartition(trainCsv$classe, p = 3/4)[[1]]
trainData <- trainCsv[inTrain,]
testData <- trainCsv[-inTrain,]
```






## GitHub Documents

This is an R Markdown format used for publishing markdown documents to GitHub. When you click the **Knit** button all R code chunks are run and a markdown file (.md) suitable for publishing to GitHub is generated.

## Including Code

You can include R code in the document as follows:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
